A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-03-07 paper/bare_conf.tex
Start perf evaluation section. Add francis' table

% TODO: ok?
\usepackage{tabulary}

@@

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only through an in-depth analysis of the data space, rather than by reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation). BLAS routines could be employed as well, although fairly complicated code manipulation would be required, for a performance gain potentially far from the expectation. The major issue would concern automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of control- and data-flow. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. In addition, register locality would be affected, since the same matrix could appear in a number of products. Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$ and on a 3D mesh, have showed run-times worse than those achieved with our optimization strategy.

@@

Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance. Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation. It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.  We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors. This choice also simplifies the compiler's cost model. The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying any other unroll-based optimizations.

@@

%TODO: NOT TRUE that kj == jk. You can have less matrices depending on j than those depending on k, meaning that less loads are needed along the j dimension.

\begin{algorithm}[t]
\caption{The PyOP2 compiler.}
\label{algo:PyOP2Compiler}
  \textbf{The PyOP2 Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper)
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
\If{isa.backend == gpu}{
  \If{it\_space}{ast.extract\_iteration\_space(wrapper)}
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
// Optimize ast based on plan \\
ast.apply\_licm() \\
ast.padding() \\
ast.data\_align() \\
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  ast.vr\_tile(plan.uaj\_factor)
}
\Return{wrapper + ast.from\_ast\_to\_c()}
\end{algorithm}

@@

\clearpage

\section{Performance Evaluation}
\label{sec:perf-results}
%Performance evaluation. We show all of the results got so far and we explain them. We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance. Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized.pdf}
\label{fig_first_case}}
\caption{Helmholtz}
\end{figure}

>>>>

% TODO: ok?

\usepackage{tabularx}
\usepackage{tabulary}

@@

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the structure and the exceptionally small size of the kernels, effective SIMD vectorization can be obtained only through an in-depth analysis of both data and iteration space, which is not supported by state-of-the-art polyhedral compilers. BLAS routines could be employed as well, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplications. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. As detailed in Section~\ref{sec:perf-results}, hand-made BLAS implementations of the Helmholtz assembly kernel (illustrated later) have run-times worse than those achieved with our optimization strategy.

%(our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation)

@@

Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance. Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation. It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.  We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors. This choice also simplifies the compiler's cost model. The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying loop unrolling, even if specific pragmas are added.

@@

\begin{algorithm}[t]
\caption{The PyOP2 compiler.}
\label{algo:PyOP2Compiler}
  \textbf{The PyOP2 Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper)
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
\If{isa.backend == gpu}{
  \If{it\_space}{ast.extract\_iteration\_space(wrapper)}
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
// Optimize ast based on plan \\
ast.licm() \\
ast.padding() \\
ast.data\_align() \\
\If{plan.permute}{
  ast.permute\_assembly\_loops()
}
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  ast.op\_vect(plan.uaj\_factor)
}
\Return{wrapper + ast.from\_ast\_to\_c()}
\end{algorithm}

@@

\section{Performance Evaluation}
\label{sec:perf-results}

Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi. The \texttt{icc 2013} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems expressible with Firedrake:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion the ``Diffusion'' equation is considered, which uses the same differential operators as Helmholtz. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$. Burgers is a time-dependent problem (i.e. the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz and Advection-Diffusion. The assembly code, partly exposed in Figure~\ref{code:burgers}, is therefore fairly different, with a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) employed.

These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method. Intuitively, the larger the element shape, the bigger is the iteration space. Triangles (2D), tetrahedron (3D), and prisms (3D) were tested. For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$. Moving from 2D to 3D also increases the number of basis function arrays, since conceptually the behaviour of the equation has to be approximated also along the $z$ dimension. On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays). A range of polynomial orders, from $p=1$ to $p=4$, are tested; higher polynomial orders are excluded from the study because of current Firedrake limitations. In such a large space of problems, the size of the stiffness matrix rarely excees 20$\times$20, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.

For the Helmholtz 3D problem, manual implementations based on MKL BLAS were tested on Sandy Bridge. This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplications that can be computed via calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used. As anticipated, both extraction of matrix-matrix multiplications from analysis of the kernel's AST and invasive re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work. However, these experiments justify that there's a set of problems for which turning to BLAS is not beneficial in terms of performance. It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might inhibit the performance gain.

Table~\ref{table:perf-results}
%In addition, register locality would be affected, since the same matrix could appear in a number of products.

%The problem is time-dependent, so , which intuitively increaes the computation cost.

%Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?

%\begin{table}[h]
%\begin{tabulary}{1.0\columnwidth}{|C|C|}
%\hline
%Problem & shape & p & original & $licm$ & $licm$ \\\hline\hline
%Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
%Padding &  \\ \hline
%Data Alignment & \\ \hline
%Loop interchange      & loops  \\ \hline
%Loop unrolling  & unroll factor \\ \hline
%Register tiling & tile size \\ \hline
%Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
%Assembly splitting ($split$) & split point, split factor \\ \hline
%\end{tabulary}
%\caption{.}
%\label{table:perf-results}
%\end{table}

%\newcommand{\massmatrixtwodresultsnorms}{
%\begin{tabularx}{1.0\textwidth}{c|XX|XX|XX|XX|}
%\cline{2-9}
%& \multicolumn{2}{c|}{$n_f=1$} & \multicolumn{2}{c|}{$n_f=2$} & \multicolumn{2}{c|}{$n_f=3$} & \multicolumn{2}{c|}{$n_f=4$} \\
%\cline{2-9}
%& A & Tensor& A & Tensor& A & Tensor& A & Tensor\\
%\hline
%\multicolumn{1}{|c|}{$p=1,q=1$} & $6.89e{-}17$ & $9.30e{-}17$ & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
%\multicolumn{1}{|c|}{$p=1,q=2$} & $6.80e{-}17$ & $1.56e{-}16$ & $1.22e{-}16$ & $7.14e{-}17$ & $1.17e{-}17$ & $1.27e{-}17$ & $8.05e{-}18$ & $\mathbf{9.77e{-}5}$ \\
%\multicolumn{1}{|c|}{$p=1,q=3$} & $8.16e{-}17$ & $8.94e{-}17$ & $6.21e{-}17$ & $\mathbf{6.77e{-}5}$ & $2.44e{-}17$ & $\mathbf{7.69e{-}4}$ & $8.61e{-}17$ & $\mathbf{5.02e{-}4}$ \\
%\multicolumn{1}{|c|}{$p=1,q=4$} & $3.70e{-}16$ & $\mathbf{1.83e{-}4}$ & $7.12e{-}16$ & $\mathbf{4.11e{-}4}$ & $2.65e{-}16$ & $\mathbf{5.42e{-}4}$ & $5.03e{-}16$ & $\mathbf{4.36e{-}4}$ \\
%\hline
%\multicolumn{1}{|c|}{$p=2,q=1$} & $2.42e{-}16$ & $2.34e{-}16$ & $1.18e{-}16$ & $3.42e{-}17$ & $2.78e{-}17$ & $\mathbf{1.30e{-}4}$ & $1.22e{-}16$ & $\mathbf{3.06e{-}4}$ \\
%\multicolumn{1}{|c|}{$p=2,q=2$} & $4.50e{-}16$ & $1.47e{-}16$ & $7.47e{-}17$ & $\mathbf{7.04e{-}5}$ & $4.05e{-}16$ & $\mathbf{6.47e{-}4}$ & - & - \\
%\multicolumn{1}{|c|}{$p=2,q=3$} & $1.85e{-}16$ & $\mathbf{1.65e{-}4}$ & $7.76e{-}16$ & $\mathbf{5.57e{-}4}$ & $1.88e{-}15$ & $\mathbf{1.47e{-}3}$ & - & - \\
%\multicolumn{1}{|c|}{$p=2,q=4$} & $2.22e{-}15$ & $\mathbf{2.90e{-}4}$ & $4.01e{-}15$ & $\mathbf{1.42e{-}3}$ & - & - & - & - \\
%\hline
%\multicolumn{1}{|c|}{$p=3,q=1$} & $6.54e{-}17$ & $5.96e{-}17$ & $8.13e{-}17$ & $\mathbf{9.70e{-}5}$ & - & - & - & - \\
%\multicolumn{1}{|c|}{$p=3,q=2$} & $9.43e{-}17$ & $1.26e{-}16$ & $8.99e{-}16$ & $\mathbf{5.23e{-}4}$ & - & - & - & - \\
%\multicolumn{1}{|c|}{$p=3,q=3$} & $2.27e{-}16$ & $\mathbf{2.86e{-}4}$ & $5.77e{-}15$ & $\mathbf{1.81e{-}3}$ & - & - & - & - \\
%\multicolumn{1}{|c|}{$p=3,q=4$} & $2.45e{-}15$ & $\mathbf{5.31e{-}4}$ & $1.62e{-}14$ & $\mathbf{2.54e{-}3}$ & - & - & - & - \\
%\hline
%\end{tabularx}
%}
%\begin{figure*}
%\massmatrixtwodresultsnorms
%\end{figure*}

\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized.pdf}
\label{fig_first_case}}
\caption{Helmholtz}
\end{figure}
